Fusing Filters with Integer Linear Programming

### 2014-05-07 2014betterfusionforfilters/section/04-ILP.tex
ilp

\subsubsection{The simple version}
\TODO{Why is this ``the simple version''? Why present a simple version before the real one?} \TODO{We need to show a complete constraint system at the start. Maybe use a simple objective function to start with, then refine it in a later section.}

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
           \> \ldots
\end{tabbing}
The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators. \ben{What is a subrate? How do we determine the subrates of each rate?}
\TODO{Expand}.

@@

\subsection{Subtype optimisations}
For two clusters $i, j$, if $x_{ij} = 0 \vee \tau_i \not= \tau_j$, then $i$'s generator must be fused with $j$ and vice versa, recursively ($x_{ij} = 0 \implies check(i,j)$).
Problem: actually generating constraints for \emph{all} pairs of nodes with unequal types is excessive.
I see two solutions to this:
\begin{itemize}
\item
Generate fewer constraints and rely on transitivity.
This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.
\item
Alternatively:
we can say, only generate constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
Here, we generate fewer constraints than originally, but still more than above.
We also must preprocess the graph to find such blocking paths.
But we still get the minimisation to count non-connected nodes.
\end{itemize}

@@

\subsubsection{An optimised version}
\label{s:OptimisedConstraints}
We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
Likewise with clusters of different, incompatible types.
The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.

@@

\subsection{Better than stream fusion}
Abdallah suggests something like:
\[
\forall g.\ \exists c.\ ilp(g, c) \le_c streamfusion(g)
\]
ie, for any graph there exists a criterion to optimise against, where we are better than or equal to stream fusion.

\subsection{Cost function}
We argue that our cost function makes intuitive sense, by reducing the number of manifest arrays and loops.
However, due to the nature of our heuristic cost function we have no strong guarantees that the clustering with the lowest cost will actually be the clustering with the shortest runtime.
In general, since we do not know the expected sizes of the arrays, this is impossible to guarantee correctly.
It is possible that for certain restricted subsets of the language, a more sophisticated cost function may be able to guarantee optimality.
This is left to future work.

@@

Now, prove that this assignment actually satisfies the constraints.

>>>>

\subsubsection{Unoptimised version}
Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent.

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_i=\emptyset \wedge trans_j=\emptyset$)                 \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge trans_j=\{j'\}$) \\
           \> \ldots
\end{tabbing}
The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators. \ben{What is a subrate? How do we determine the subrates of each rate?}
\TODO{Expand}.

@@

\subsection{A note on transitivity}
It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.

\subsubsection{Fusion-preventing path optimisation}
\label{s:OptimisedConstraints}
Alternatively:
we can say, only generate constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
Here, we generate fewer constraints than originally, but still more than above.
We also must preprocess the graph to find such blocking paths.
But we still get the minimisation to count non-connected nodes.

@@

We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
Likewise with clusters of different, incompatible types.
The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.

@@

@@

\TODO{Now, prove that this assignment actually satisfies the constraints.}
